---
title: "Moving Beyond Tree Variable Importance: Bagged Trees"
author: "Aurora Owens"
date: "December 1st, 2016"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    highlight: kate
    theme: spacelab
---

```{r loadlib, message=FALSE, warning=FALSE}
library(MASS)
library(ggplot2)
library(dplyr)
library(tree)
library(grid)
library(gridExtra)
library(ipred)
```

```{r multiplot}
# Multiple plot function
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}
```

```{r functionsVIM}
intervalReturn <- function(split.cutleft){ 
  intervals <- data.frame(split.cutleft)
  intervals <- filter(intervals, split.cutleft != "")
  intervals <- dplyr::select(intervals, split.cutleft)
  intervals$split.cutleft <- gsub("<", "", intervals$split.cutleft)
  intervals$split.cutleft <- as.numeric(intervals$split.cutleft)
  return(arrange(intervals, (split.cutleft)))
}

partitionCreator <- function(i = interval, x, data) {
  p <- c(1:(1+nrow(i)))
  bins <- data.frame (bin = rep(1, nrow(data)), x = data[,x])
  for(j in 1:nrow(i)){
    bins[bins$x > i[j,1],1] <- p[j+ 1]
  }
  return(bins$bin)
}

condPermuter <- function(d2, x) {
  d2$p <- as.factor(d2$p)
  for(i in 1:length(levels(d2$p))){
   #set.seed(1)
   d2[d2$p == i, x] <- sample(d2[d2$p== i,x])
  }
  return(d2[,x])
}

VIdevBagged<- function(b, x){
  return(as.numeric(as.data.frame(b$mtrees[[1]]$btree$frame)[,c(1,4)] %>% filter(var == x) %>% dplyr::summarise(sum = sum(dev))))
}

variable.importance.dev <- function(t1, x, d){
  if(length(t1$frame$splits[,1]) == 1) {
    stop(paste(x, "is not correlated strongly with the other predictors, try another variable importance procedure"))
  } else {
  interval <- intervalReturn(t1$frame$splits[,1])
  d$p <- partitionCreator(interval, x, d)
  d[,x] <- condPermuter(d, x)
  t2 <- tree(y ~ ., data = d, y = TRUE)
  v1 <- VIdevBagged(t2, x)
  #rm(t1, t2)
  return(v1)
  }
}
```


```{r getData}

#same process as in `sampling.Rmd`
rep(0, 12) -> mu #mean of each variable

diag(12) -> sigma #creates a 12 by 12 diagonal matrix with 1's down the diagonal

sigma -> sigma1
.9 -> sigma1[1,2]
.9 -> sigma1[2,1] #adding the block correlation between the first four variables
.9 -> sigma1[1,3]
.9 -> sigma1[3,1]
.9 -> sigma1[3,2]
.9 -> sigma1[2,3]
.9 -> sigma1[1,4]
.9 -> sigma1[4,1]
.9 -> sigma1[2,4]
.9 -> sigma1[4,2]
.9 -> sigma1[4,3]
.9 -> sigma1[3,4]

#sigma1

1000 -> n #number of observations
set.seed(1)
mvrnorm(n =n, mu = mu, Sigma = sigma1) -> sim1000 #sampling from the multivariate normal 

c(5,5,2,0,-5,-5,-2,0,0,0,0,0) -> bts #these are the betas 

rnorm(1000, mean = 0, sd = .5) -> e #error terms

rep(0, 1000) -> ys #init a vector of zeros

for( i in 1:1000){ #go row by row and create the ys based on the function of e, bts, and sim1000
ys[i] <- sim1000[i,1]*bts[1]+
    sim1000[i,2]*bts[2]+ 
    sim1000[i,3]*bts[3]+ 
    sim1000[i,4]*bts[4]+ 
    sim1000[i,5]*bts[5]+ 
    sim1000[i,6]*bts[6]+
    sim1000[i,7]*bts[7]+ 
    sim1000[i,8]*bts[8]+ 
    sim1000[i,9]*bts[9]+ 
    sim1000[i,10]*bts[10]+ 
    sim1000[i,11]*bts[11]+ 
    sim1000[i,12]*bts[12] +
    e[i]
}

mvn.data <- as.data.frame(sim1000)
mvn.data$y <- ys
```

This report will use the same data as the previous files, so that there are 12 predictors where the first four are block correlated to each other with $\rho = .9$. They are related to $Y$ by the linear equation: $$Y = 5 \cdot V_1 + 5 \cdot V_2 + 2 \cdot V_3 + 0 \cdot V_4 + -5 \cdot V_5 + -5\cdot V_6 + 0\cdot V_7 + 0 \cdot ..... + E$$ Where the coefficents for variables 7-12 are all zero and $E$ ~ $N(0,\frac 1 2 )$. 

The initial tree of the model $Y$ ~$X_1, X_2, ..., X_{12}$ in `another_vim.Rmd` was very variable. To get a more stable initial variable importance, we'll introduce a bagged model. 

```{r}
b <- bagging(y ~., mvn.data)
#btree <- as.data.frame(b$mtrees[[1]]$btree$frame)

v0 <- list(rep(0,25), rep(0,25),rep(0,25),rep(0,25),rep(0,25),rep(0,25),
           rep(0,25),rep(0,25),rep(0,25),rep(0,25),rep(0,25),rep(0,25))

vs <- names(mvn.data[,-13])

for(j in 1:12){
  v <- rep(0,25)
  for(i in 1:25){
   as.numeric(as.data.frame(b$mtrees[[i]]$btree$frame)[,c(1,4)] %>% 
                filter(var == vs[j]) %>% 
                dplyr::summarise(sum = sum(dev))) -> v[i]
  }
  v0[[j]] <- (v)
  rm(v) 
}
```
**Non-Permuted Variable Importance of the first 10 Bagged Trees**

|      Bagged Tree      |     V1     |      V2      |     V3     |    V4    |     V5    |    V6    | V7 | V8 | V9 | V10 | V11 | V12 |
|:---------------------:|:----------:|:------------:|:----------:|:--------:|:---------:|:--------:|:--:|:--:|:--:|:---:|:---:|:---:|
|           1           |  203726.09 |  117195.788  | 14420.756  |     0    | 36013.22  | 48139.37 |  0 |  0 |  0 |  0  |  0  |  0  |
|           2           |  37875.12  |  278403.597  |      0     |     0    | 42828.19  | 38683.29 |  0 |  0 |  0 |  0  |  0  |  0  |
|           3           |  312004.01 |      0       |  5642.170  | 6370.531 | 60627.11  | 22928.16 |  0 |  0 |  0 |  0  |  0  |  0  |
|           4           |  71653.62  |  245228.504  |  22561.345 |     0    | 48500.97  | 29829.11 |  0 |  0 |  0 |  0  |  0  |  0  |
|           5           | 240861.08  |   50524.628  |  6535.479  |     0    | 36298.52  | 51793.57 |  0 |  0 |  0 |  0  |  0  |  0  |
|           6           | 108356.37  |  220651.893  |     0      | 7362.526 | 48477.18  | 48179.97 |  0 |  0 |  0 |  0  |  0  |  0  |
|           7           | 254026.51  |       0      |  55727.322 |     0    | 44607.51  | 37730.76 |  0 |  0 |  0 |  0  |  0  |  0  |
|           8           | 127419.42  |  216383.575  |      0     |     0    | 25382.48  | 41084.73 |  0 |  0 |  0 |  0  |  0  |  0  |
|           9           | 319015.03  |   10417.410  | 15760.648  |     0    | 34298.68  | 33702.96 |  0 |  0 |  0 |  0  |  0  |  0  |
|           10          |  68920.82  |  220037.046  |  55502.950 |     0    |  53128.59 | 45627.94 |  0 |  0 |  0 |  0  |  0  |  0  |
| Original Coefficients |      5     |       5      |      2     |     0    |     -5    |    -5    | 0 |  0 |  0 |  0  |  0  |  0  |

