---
title: "Simulations"
author: "Aurora Owens"
date: "October 3rd, 2016"
output:
  html_document:
    highlight: kate
    theme: spacelab
---

```{r loadlib}
library(ggplot2)
library(dplyr)
```
 
#A Very Simple Tree and a Measure of Variable Importance

##Data Simulations

A Three Dimensional Example Where $X_1$ and $X_2$ Are Non-Linearly Correlated

```{r}

set.seed(1)

x1 <- rnorm(1000)

set.seed(2)

x2 <- exp(x1) + rnorm(100)

#x2 <- 5* sin(3*x1) + 2*cos(3*x1+ .5) + rnorm(1000)

set.seed(3)

y <- 5*x1 + 4*x2 + rnorm(100)
```

We can see by plotting x1 and x2 that their relationship is not linear, but their relationship does still have some positive linear correlation. 

```{r, echo = FALSE}
data <- data.frame("y" = y, "x1" = x1, "x2" = x2)
ggplot(aes(x = x1, y = x2), data = data) + 
  geom_point()+ 
  ggtitle("Relation of X1 and X2")

cor(x1,x2)
```


Their linear correlation is .711. 

```{r, echo=FALSE}
ggplot(aes(x = x1, y = y), data = data) +
  geom_point()+
  ggtitle("X1 vs Y")

ggplot(aes(x = x2, y = y), data = data)+
  geom_point()+
  ggtitle("X2 vs Y")
```

Y appears to be correlated with X1 but that is only because they are both so highly correlated with X2. Let's see how this pattern effects a simple tree. 

#A new Variable Importance Measure

```{r}
intervalReturn <- function(split.cutleft){ 
  intervals <- data.frame(split.cutleft)
  intervals <- filter(intervals, split.cutleft != "")
  intervals <- select(intervals, split.cutleft)
  intervals$split.cutleft <- gsub("<", "", intervals$split.cutleft)
  intervals$split.cutleft <- as.numeric(intervals$split.cutleft)
  return(arrange(intervals, (split.cutleft)))
}
```


```{r}
library(tree)
library(randomForest)
set.seed(428)
rfo <- randomForest(y~., data = data[,-4])
importance(rfo)


set.seed(222)
tree1 <- tree(x2~x1, data = data)
plot(tree1)
text(tree1)

tree1$frame
#Partition X2 wrt the splits on X1
interval <- intervalReturn(tree1$frame$splits[,1])

#This code should be written as a function to handle more cases 
data$partition.x2 <- ifelse(x1 < interval[1,1], 1, 
                         ifelse(x1 < interval[2,1], 2, 
                         ifelse(x1 < interval[3,1], 3,
                         ifelse(x1 < interval[4,1], 4,
                         ifelse(x1 < interval[5,1], 5,
                         ifelse(x1 < interval[6,1], 6, 7))))))


#Permute X2 wrt these partitions
data.permuted.x2 <- data

d1 <- data.permuted.x2[data.permuted.x2$partition.x2 == 1,] 
d2 <- data.permuted.x2[data.permuted.x2$partition.x2 == 2,] 
d3 <- data.permuted.x2[data.permuted.x2$partition.x2 == 3,] 
d4 <- data.permuted.x2[data.permuted.x2$partition.x2 == 4,] 
d5 <- data.permuted.x2[data.permuted.x2$partition.x2 == 5,] 
d6 <- data.permuted.x2[data.permuted.x2$partition.x2 == 6,] 
d7 <- data.permuted.x2[data.permuted.x2$partition.x2 == 7,]

d1$x2 <- sample(d1$x2)
d2$x2 <- sample(d2$x2)
d3$x2 <- sample(d3$x2)
d4$x2 <- sample(d4$x2)
d5$x2 <- sample(d5$x2)
d6$x2 <- sample(d6$x2)
d7$x2 <- sample(d7$x2)

data.permuted.x2 <- rbind(d1,d2,d3,d4,d5,d6,d7)

#Fit a new rf
set.seed(1)
rf2 <- randomForest(y~., data = data.permuted.x2[,-4])
importance(rf2)

#Do the same with X1
set.seed(222)
tree2 <- tree(x1~x2, data = data)
plot(tree2)
text(tree2)

#tree1$frame
#Partition X1 wrt the splits on X2
interval1 <- intervalReturn(tree2$frame$splits[,1])

#This code should be written as a function to handle more cases 
data$partition.x1 <- ifelse(x2 < interval1[1,1], 1, 
                         ifelse(x2 < interval1[2,1], 2, 
                         ifelse(x2 < interval1[3,1], 3,
                         ifelse(x2 < interval1[4,1], 4,
                         ifelse(x2 < interval1[5,1], 5,
                         ifelse(x2 < interval1[6,1], 6, 7))))))


#Permute X1 wrt these partitions
data.permuted.x1 <- data

d1 <- data.permuted.x1[data.permuted.x1$partition.x1 == 1,] 
d2 <- data.permuted.x1[data.permuted.x1$partition.x1 == 2,] 
d3 <- data.permuted.x1[data.permuted.x1$partition.x1 == 3,] 
d4 <- data.permuted.x1[data.permuted.x1$partition.x1 == 4,] 
d5 <- data.permuted.x1[data.permuted.x1$partition.x1 == 5,] 
d6 <- data.permuted.x1[data.permuted.x1$partition.x1 == 6,] 
d7 <- data.permuted.x1[data.permuted.x1$partition.x1 == 7,]

d1$x1 <- sample(d1$x1)
d2$x1 <- sample(d2$x1)
d3$x1 <- sample(d3$x1)
d4$x1 <- sample(d4$x1)
d5$x1 <- sample(d5$x1)
d6$x1 <- sample(d6$x1)
d7$x1 <- sample(d7$x1)

data.permuted.x1 <- rbind(d1,d2,d3,d4,d5,d6,d7)
#Fit a new rf
set.seed(1)
rf1 <- randomForest(y~., data = data.permuted.x1[,c(-4, -5)])
importance(rf1)
```


