---
header-includes:
- \usepackage{amssymb,amsthm,amsmath}
- \usepackage{chemarr}
output: pdf_document
---

<!--
You can delete the header-includes (lines 3-5 above) if you like and also the chunk below since it is loaded in the skeleton.Rmd file.  They are included so that chap2.Rmd will compile by itself when you hit Knit PDF.
-->

```{r include_reedtemplates_2, include = FALSE}
# This chunk ensures that the reedtemplates package is installed and loaded
# This reedtemplates package includes the template files for the thesis and also
# two functions used for labeling and referencing
if(!require(devtools))
  install.packages("devtools", repos = "http://cran.rstudio.com")

if(!require(reedtemplates)){
  library(devtools)
  devtools::install_github("ismayc/reedtemplates")
  }
library(reedtemplates)
library(MASS)
```
#Simulations and Comparisons

##Simulated Data

- From Strobl et al:

&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;To aid in comparisons between the methods, one of the simulated datasets considered in this paper will be generated from the same method as used in (Strobl et al, 2008???). Under this method, the 13 x 1000 data set, $D_1$, has 12 predictors, $V_1,..,V_{12}$, where $V_j \sim N(0,1)$. The first four are, however, block correlated to each other with $\rho = .9$. They are related to $Y$ by the linear equation: $$Y = 5 \cdot V_1 + 5 \cdot V_2 + 2 \cdot V_3 + 0 \cdot V_4 + -5 \cdot V_5 + -5\cdot V_6 + 0\cdot V_7 + 0 \cdot ..... + E, E \sim N(0,\frac 1 2 )$$ Note that the coefficents for $V_7,...,V_{12}$ are all zero. 

```{r stroblSim, echo=FALSE}
rep(0, 12) -> mu #mean of each variable

diag(12) -> sigma #creates a 12 by 12 diagonal matrix with 1's down the diagonal

sigma -> sigma1
.9 -> sigma1[1,2]
.9 -> sigma1[2,1] #adding the block correlation between the first four variables
.9 -> sigma1[1,3]
.9 -> sigma1[3,1]
.9 -> sigma1[3,2]
.9 -> sigma1[2,3]
.9 -> sigma1[1,4]
.9 -> sigma1[4,1]
.9 -> sigma1[2,4]
.9 -> sigma1[4,2]
.9 -> sigma1[4,3]
.9 -> sigma1[3,4]

1000 -> n #number of observations
set.seed(1)
mvrnorm(n =n, mu = mu, Sigma = sigma1) -> sim1000 #sampling from the multivariate normal 
c(5,5,2,0,-5,-5,-2,0,0,0,0,0) -> bts #these are the betas 

rnorm(1000, mean = 0, sd = .5) -> e #error terms

rep(0, 1000) -> ys #init a vector of zeros

for( i in 1:1000){ #go row by row and create the ys based on the function of e, bts, and sim1000
ys[i] <- sim1000[i,1]*bts[1]+
    sim1000[i,2]*bts[2]+ 
    sim1000[i,3]*bts[3]+ 
    sim1000[i,4]*bts[4]+ 
    sim1000[i,5]*bts[5]+ 
    sim1000[i,6]*bts[6]+
    sim1000[i,7]*bts[7]+ 
    sim1000[i,8]*bts[8]+ 
    sim1000[i,9]*bts[9]+ 
    sim1000[i,10]*bts[10]+ 
    sim1000[i,11]*bts[11]+ 
    sim1000[i,12]*bts[12] +
    e[i]
}

mvn.data <- as.data.frame(sim1000)
mvn.data$y <- ys
```

```{r stroblSimPlot, echo=FALSE, warning=FALSE}
plot(mvn.data)
```

Let's move on to a more difficult situation. The dataset $D2$ contains five predictors, $X_1,...X_5$, that have an interesting structure- several of the predictors are correlated, but are not one -to -one. This violates an important assumtion of the linear model and means that these variables have low correlation. Note that this only makes sense in higher dimensions where we are estimating the value of $X_j$ given $X_1,...,X_n$. 

```{r, sqrtAbs}
x1 <- rnorm(1000)

x2 <- 2*sqrt(abs(x1)) + rnorm(1000)

x3 <- x1 + 2*x2 + rnorm(1000)

x4 <- rnorm(1000)

x5 <- 2*sqrt(abs(x4)) + rnorm(1000)

y <- x1 + 2*x2 + 3 * x3 + 4*x4 + 1*x5 + rnorm(1000)

d2 <- data.frame(y,x1,x2,x3,x4,x5)

cor(d2)

plot(d2)
```

The tricker relationship between the variables in $D2$ was because they were generated using the square root of the absolute value of the other variable. Here, in $D3$ the process is repeated but with the log of the absolute value. 

```{r logAbs}
w1 <- rnorm(1000)

w2 <- 2*log(abs(w1)) + rnorm(1000)

w3 <- w1 + 2*w2 + rnorm(1000)

w4 <- rnorm(1000)

w5 <- 2*log(abs(w4)) + rnorm(1000)

y <- w1 + 2*w2 + 3 * w3 + 4*w4 + 1*w5 + rnorm(1000)

d3 <- data.frame(y,w1,w2,w3,w4,w5)

cor(d3)

plot(d3)
```

##Models and Comparisons
##Trees
###CART
  In 1984, Breiman et al introduces a revolutionary new algorithm for trees. **Need to acquire** _Classification and Regression Trees_  **to make sure the method discused in MASS is the same that Breiman uses/is used in** `randomForest`

**Tree Algorithm** CART?
 
Begin by considering the entire feature space $X_1, ..., X_n$. Then:

1. Consider every possible pair of partitions of this feature space, $P_1, P_2$, so that if $X_1 = x_1 , X_2 = x_2,..., X_n = x_n$ where ${x_1,...,x_n}  \in P_1$ then our prediction is the mean value of $y$ given $x_1,..,x_n \in P_1$. 

2. Choose the partitions that minimize RSS 

3. For each new partition, repeat steps 1 and 2 until some stopping condition is reached. 

###Binary Recursive Partitioning 

###Conditional Inference Trees
CI trees
1. For case weights  $w$  test the global null hypothesis of independence between any of the m covariates and the response. Stop if this hypothesis cannot be rejected. Otherwise select the  $j_{th}$  covariate  $Xj$  with strongest association to  $Y$.

2. Choose a set  $A \subset X_{j}$  in order to split  $X_{j}$  into two disjoint sets  $A$  and  $X_{j}$ \ $A$.  The case weights  $w_{left}$  and  $w_{right}$  determine the two subgroups with  $w_{left,i} = w_iI(X_{j\cdot i} \in A)$  and  $w_{right,i} = w_iI(X_{ji} \in A)$ for all $i = 1,...,n$  ( $I(·)$  denotes the indicator function). 

3. Recursively repeat steps 1 and 2 with modified case weights  $w_left$  and  $w_right$, respectively. 

 from  https://eeecon.uibk.ac.at/~zeileis/papers/Hothorn+Hornik+Zeileis-2006.pdf

After step 1 is completed, any goodness of fit method can be used to generate the split and choose the set $A$. Note that in this method the splitting is done separately from the variable selection. 

##Bagged Forests

##Random Forests




